---
title: "HashtagISME16"
author: "Keith H. Turner (keith.turner@monsanto.com, http://twitter.com/kay_aych)"
date: "8/27/2016"
output: html_document
runtime: shiny
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = F, error = F, message = F)
```

*__Update 8/31/2016:__ Added in some graph clustering analysis (at the bottom). More importantly, properly captioned the moose gif.*

*The code and data for this post are up on [my GitHub](https://github.com/khturner/HashtagISME16), please feel free to fork it, or drop me a line if you have any comments/questions.*

ISME16 in Montreal was a blast. There were so many great talks and posters, and lots of cool, friendly people. As with many academic conferences these days, we were very active on Twitter too, with the hashtag [#ISME16](https://twitter.com/hashtag/ISME16?src=hash). We basically took Montreal Twitter over at the start of the conference:

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr"><a href="https://twitter.com/hashtag/ISME16?src=hash">#ISME16</a> is trending topic in <a href="https://twitter.com/Montreal">@Montreal</a> ????</p>&mdash; The ISME Office (@ISME_OFFICE) <a href="https://twitter.com/ISME_OFFICE/status/767468449769676800">August 21, 2016</a></blockquote>
<script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>

### Downloading tweets

Over 6,800 tweets were posted with the #ISME16 hashtag, which should be a large enough dataset to do some fun stuff with. First though, let's download them.

```{r, eval=F}
library(dplyr)
library(jsonlite)
library(twitteR)

# Set up Twitter access - I saved my app API credentials in a JSON
twitter_credentials <- readChar("~/twitter_credentials.json", 1000) %>% fromJSON
setup_twitter_oauth(twitter_credentials$consumer_key,
                    twitter_credentials$consumer_secret,
                    twitter_credentials$access_token,
                    twitter_credentials$access_secret)

# Get tweets!
isme16_tweets <- searchTwitter("#ISME16", n = 10000) %>% twListToDF() %>%
  tbl_df %>% mutate(created = created - 4*60*60) # Montreal is 4 hours back 

# Save results
saveRDS(isme16_tweets, "isme16_tweets.rds")
```

```{r, echo=F, warning=F, error=F}
library(dplyr)
```

In this document, I'm not running the above code to avoid hitting the Twitter API every time it loads, so let's load the saved results and take a look at them.

```{r}
isme16_tweets <- readRDS("isme16_tweets.rds")

# Let's take a look
isme16_tweets

```

### Activity over time

There's a lot of metadata associated with these tweets that we'll get to in a bit, but first let's just look at the most simple one - when a tweet was posted. When were the most active times during ISME16?

```{r fig.width=12}
library(ggplot2)
library(RColorBrewer)
tweets_week_of <- isme16_tweets %>%
  filter(created >= "2016-08-21") # Let's ignore pre-ISME tweets for now
tweets_week_of %>%
  ggplot(aes(created, fill = isRetweet)) + geom_histogram(bins = 100) + 
  scale_x_datetime(date_breaks = "6 hours", date_labels = "%a %d %H:%M") +
  theme_bw() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  scale_fill_brewer(type = "qual", palette = "Set1")
```

I think this plot tells you a lot about the week. We kicked off on **Sunday** night with a rousing, hockey stick time-enforced opening ceremony and keynote lecture by Charles Greer on hydrocarbon utilization in cold ocean microbes, and there was much tweeting.

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Charles Greer finds oil degradation capacity under arctic ice (&amp; threatened with hockey stick if over time). <a href="https://twitter.com/hashtag/ISME16?src=hash">#ISME16</a> <a href="https://t.co/zCf4DKT3bX">pic.twitter.com/zCf4DKT3bX</a></p>&mdash; Nature Microbiology (@NatureMicrobiol) <a href="https://twitter.com/NatureMicrobiol/status/767477455506534400">August 21, 2016</a></blockquote>
<script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Charles Greer on sampling ice cores in the Arctic in winter: &quot;you hope the polar bears are asleep because it&#39;s 24 hour darkness&quot; <a href="https://twitter.com/hashtag/ISME16?src=hash">#ISME16</a></p>&mdash; Laura Hug (@LAHug_) <a href="https://twitter.com/LAHug_/status/767465001078689792">August 21, 2016</a></blockquote>
<script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>

After that the conference got started in earnest. You can tell from the tweets by time plot that there were sessions after 17:00 on **Monday** (the roundtables):

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Soil metagenomics round table: so many analysis tools, but which/how to use and when??? <a href="https://twitter.com/hashtag/ISME16?src=hash">#ISME16</a> <a href="https://t.co/FlCt8ZNKhu">pic.twitter.com/FlCt8ZNKhu</a></p>&mdash; Nature Microbiology (@NatureMicrobiol) <a href="https://twitter.com/NatureMicrobiol/status/767865190322835456">August 22, 2016</a></blockquote>
<script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>


and **Thursday** (Penny Chisolm's wonderful Tiedje Award talk on her career in *Prochlorococcus*):

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Penny Chisholm <a href="https://twitter.com/ChisholmLab_MIT">@ChisholmLab_MIT</a> receives Tiedje Award for her pioneering work in Prochlorococcus. Congrats! <a href="https://twitter.com/hashtag/ISME16?src=hash">#ISME16</a> <a href="https://t.co/lUqBOAvE8N">pic.twitter.com/lUqBOAvE8N</a></p>&mdash; Nature Microbiology (@NatureMicrobiol) <a href="https://twitter.com/NatureMicrobiol/status/768957228535648257">August 25, 2016</a></blockquote>
<script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>

but people were too busy partying on **Tuesday** evening:

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr"><a href="https://twitter.com/hashtag/ISME16?src=hash">#ISME16</a> party update. <a href="https://t.co/oqSkuTsrTl">pic.twitter.com/oqSkuTsrTl</a></p>&mdash; Josh Neufeld (@joshdneufeld) <a href="https://twitter.com/joshdneufeld/status/768285318600531968">August 24, 2016</a></blockquote>
<script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>

and flying out on **Friday** evening:

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr"><a href="https://twitter.com/hashtag/ISME16?src=hash">#ISME16</a> was great! Made new friends and (hopefully) kept the old :) Thanks for the good times everyone, safe travels!</p>&mdash; Daan Speth (@daanspeth) <a href="https://twitter.com/daanspeth/status/769360501994291201">August 27, 2016</a></blockquote>
<script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>

to tweet much. Another thing I noticed was that **at night**, while most of us conferencegoers were resting up for the next day of science, the ratio of retweets to tweets went way up.

```{r fig.width=12}
tweets_week_of$bin <- cut(as.numeric(tweets_week_of$created), breaks = 100)
tweets_week_of %>% group_by(bin) %>%
  summarize(tweets = n(),
            retweetsToTweets = sum(isRetweet) / tweets,
            created = min(created)) %>%
  ggplot(aes(created, tweets, fill = retweetsToTweets)) + geom_bar(stat = "identity") +
  scale_x_datetime(date_breaks = "6 hours", date_labels = "%a %d %H:%M") +
  theme_dark() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + 
  scale_fill_distiller(palette = "Spectral")

```

One hypothesis that comes to mind is that #ISME16 tweets continued to echo around the world while Montreal slept. Is that true? Well unfortunately it's hard to tell, because so few of these tweets are geocoded.

```{r}
tweets_week_of %>% mutate(geocoded = !is.na(longitude)) %>% group_by(geocoded) %>% tally
```

Of course the most noticably aberrant period on the tweets by time plot is **Wednesday**, when we had the day off to look up from our smartphones and explore Montreal.

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Free day at <a href="https://twitter.com/hashtag/ISME16?src=hash">#ISME16</a> was so great. View from mount royale <a href="https://t.co/epQYDHNkQ2">pic.twitter.com/epQYDHNkQ2</a></p>&mdash; Erick Cardenas (@erickcardenasp) <a href="https://twitter.com/erickcardenasp/status/768598296575242240">August 24, 2016</a></blockquote>
<script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>
<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Day to explore during <a href="https://twitter.com/hashtag/ISME16?src=hash">#ISME16</a> Spent most of day at splendid &amp; huge Montreal Botanical Garden! Hope everyone had as much fun as M &amp; I did!!</p>&mdash; Richard E. Lenski (@RELenski) <a href="https://twitter.com/RELenski/status/768560509662748672">August 24, 2016</a></blockquote>
<script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>


### Most active users

I can't seem to find the official figure anywhere, but ISME says that its symposia typically attract [between 1500 and 2500 delegates](http://www.isme-microbes.org/isme16/practical), which I would say felt about right. How many people were tweeting using the #ISME16 hashtag, and who was the most active? The most favorited/retweeted?

```{r}
original_tweets_week_of <- tweets_week_of %>% filter(!isRetweet) # Only original tweets
tweeter_summary <- original_tweets_week_of %>%
  group_by(screenName) %>% # For each screenname, summarize several counts
  summarize(numTweets = n(), numFavorites = sum(favoriteCount),
            numRetweets = sum(retweetCount)) %>% arrange(-numTweets)
renderDataTable(tweeter_summary)
```

So it looks like ~350 people were active on Twitter during ISME16, which is a pretty good ratio! I'd be interested to see how that compares to ISME15, and what it'll be at ISME17. As for activity, man, I thought I tweeted a lot this week, but [Siouxsie Wiles](https://twitter.com/SiouxsieW) has all of us beat by far. What's the distribution of activity look like among the top 30 active tweeters at #ISME16 this week?

```{r}
tweeter_summary %>% slice(1:30) %>%
  transform(screenName = reorder(screenName, -numTweets)) %>% # Order descending
  ggplot(aes(screenName, numTweets)) + geom_bar(stat = "identity") +
  theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

So it looks like there were a handful of super-tweeters with a long tail of casual tweeters. We have some other metrics in these first-order data though, including the number of retweets and favorites. If we consider those as "impact", who are the 30 most impactful ISME16 tweeters? And what's their "impact ratio" (i.e. "impact" per number of tweet)?

```{r}
tweeter_summary %>% mutate(impact = numFavorites + numRetweets) %>%
  arrange(-impact) %>% slice(1:30) %>%
  transform(screenName = reorder(screenName, -impact)) %>%
  ggplot(aes(screenName, impact, fill = impact / numTweets)) +
  geom_bar(stat = "identity") +
  theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))  + 
  scale_fill_distiller(palette = "Spectral",
                       trans = "log", breaks = c(1, 5, 10, 50))
```

I think this is starting to show some interesting divisions in Twitter patterns. First, impact (as measured by favorites + retweets) seems to be a bit more even across the top accounts than activity. Second, there are several patterns on display here, showing a range in both proflicacy and per-tweet impact. Some tweeters, for example, [Thea Whitman](https://twitter.com/TheaWhitman), had big impact with just a few tweets, most notably, ISME bingo (which I believe could be won by Tuesday afternoon):

<blockquote class="twitter-tweet" data-lang="en"><p lang="en" dir="ltr">Want to play <a href="https://twitter.com/hashtag/ISME16?src=hash">#ISME16</a> bingo? <a href="https://t.co/RIk2WiLSXC">pic.twitter.com/RIk2WiLSXC</a></p>&mdash; Thea Whitman (@TheaWhitman) <a href="https://twitter.com/TheaWhitman/status/768111610628018176">August 23, 2016</a></blockquote>
<script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>

### Mentions: Twitter as a conversation platform

One of my favorite things about using Twitter at conferences though is its use as a forum for real-time conversation between participants. It has great potential to span different sessions, times, and physical boundaries. With ~2,000 attendees, you can't possibly always meet up with the right group to discuss something you're interested in. Who were some of the most active tweeters in conversations? We'll begin looking into this by compiling a record of mentions by parsing the `text` field.

```{r}
# Note: this approach does not capture conversations in the form of "quoted" retweets, as they appear as hyperlinks in the text field. This also does not count retweets as mentions.

# Make a list of mentionees, index by mentioner
mentions <- regmatches(original_tweets_week_of$text,
                       gregexpr("@[-_A-Za-z0-9]+",
                                original_tweets_week_of$text))
mentions <- lapply(mentions, function(x) gsub("@", "", x)) # Strip off @
names(mentions) <- original_tweets_week_of$screenName
# Filter out non-mentioning tweets
mentions <- mentions[lapply(mentions, length) > 0]

# Who were the 30 most mentioned tweeters?
data.frame(screenName = unlist(mentions)) %>% tbl_df %>% 
 group_by(screenName) %>%
 tally %>% arrange(-n) %>% slice(1:30) %>%
  transform(screenName = reorder(screenName, -n)) %>%
  ggplot(aes(screenName, n)) + geom_bar(stat = "identity") +
  theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1))
 
```

This list shouldn't be too surprising to anybody who was at ISME16, these were the people who gave some of the best talks of the conference (in my humble opinion!) [Rich Lenski](https://twitter.com/RELenski) captured everyone's imagination in a great keynote address discussing the dynamics of evolution in the [*E. coli* Long-Term Evolution Experiment](https://en.wikipedia.org/wiki/E._coli_long-term_evolution_experiment). [Ashley Shade](https://twitter.com/ashley17061) talked about microbial community dynamics in the [Centralia mine fire](https://en.wikipedia.org/wiki/Centralia_mine_fire) (which I didn't catch, but I have seen her talk before and loved it). And [Kelly Wrighton](https://twitter.com/kcwrighton) cut a g.d. hole in a moose:

<blockquote class="twitter-tweet" data-conversation="none" data-lang="en"><p lang="en" dir="ltr"><a href="https://twitter.com/m3thyl">@m3thyl</a> <a href="https://twitter.com/hashtag/ISME16?src=hash">#ISME16</a> K. Wrighton is a serious contender in the awesome sampling system category with a fistulated moose!! <a href="https://t.co/lRDcIGjPGU">pic.twitter.com/lRDcIGjPGU</a></p>&mdash; Adrien Assié (@m3thyl) <a href="https://twitter.com/m3thyl/status/768874608715964416">August 25, 2016</a></blockquote>
<script async src="//platform.twitter.com/widgets.js" charset="utf-8"></script>

<iframe src="//giphy.com/embed/26ufagX7zCaxYtoCQ" width="480" height="377" frameBorder="0" class="giphy-embed" allowFullScreen></iframe><p><a href="https://giphy.com/gifs/26ufagX7zCaxYtoCQ">via GIPHY</a></p>

Also, her talk about it was awesome, showing that we have a huge gap in biochemical and microbiological knowledge when it comes to the basics of ruminant digestion.

### Twitter as a social network

The unique thing about Twitter mentions as opposed to other social network tags (e.g. Facebook) is that they can mark a back-and-forth conversation. Now that we've extracted mentions from our #ISME16 tweets, we can explore the dynamics of these conversations. We'll do so with a graph, where users are nodes and mentions are edges between those nodes

```{r}

# Extract mentions as a vector of "from", "to", "from", "to"...
edge_sequence <- lapply(seq_along(mentions), function(i) {
  as.vector(rbind(rep(names(mentions)[[i]], length(mentions[[i]])),
                  mentions[[i]]))
  }) %>% unlist

# Summarize from, to and number of mentions in a df
edges <- data.frame(from = edge_sequence[seq(1, length(edge_sequence), 2)],
                    to = edge_sequence[seq(2, length(edge_sequence), 2)],
                    stringsAsFactors = F) %>% tbl_df %>%
  group_by(from, to) %>% summarize(value = n())

# Build a df for nodes
nodes <- data.frame(id = unique(c(edges$from, edges$to)),
                    label = unique(c(edges$from, edges$to)),
                    stringsAsFactors = F) %>% tbl_df

# Construct an igraph object of our mention graph
library(igraph)
mention_graph <- make_empty_graph() + vertices(nodes$id) +
  edges(as.vector(rbind(edges$from, edges$to)), value = edges$value)

# Calculate centrality of our nodes with PageRank (scaled a bit)
V(mention_graph)$value <- page_rank(mention_graph, weights = E(mention_graph)$value)[[1]] - 0.0013 

# Did a user use the #ISME16 hashtag?
V(mention_graph)$group <- ifelse(V(mention_graph)$name %in% tweeter_summary$screenName, "HashtagUser", "NonHashtagUser")

# Visualize it! In this visualization, blue nodes are people who used the #ISME16 hashtag, and yellow are those who didn't. An arrow is drawn from one node to another if the first node mentions the second in a tweet with the #ISME16 hashtag. Nodes are sized by PageRank.
library(visNetwork)
mention_graph_vn <- toVisNetworkData(mention_graph)
visNetwork(nodes = mention_graph_vn$nodes, edges = mention_graph_vn$edges,
           width = "100%", height = "800px") %>%
  visIgraphLayout(physics = T, smooth = T) %>%
  visEdges(arrows = "to") %>%
  visOptions(highlightNearest = TRUE)

```

So there's clearly a lot to chew on here, but I think there's some pretty neat stuff going on. For example, you can definitely see again that the plenary speakers (Rich Lenski, Penny Chisolm, etc.), are huge central points in the discussion. But you can also identify some people who are highly interconnected in the ISME16 sub-Twittersphere. Also interesting are those people around the edges who used the #ISME16 hashtag (and thus are blue), who "spread the gospel" of the conference to others who didn't use the hashtag (and thus are yellow).

### Clustering, cliques, and expanding your network at conferences

While browsing around in that mention graph, I noticed a few interconnected cliques buried in the whole #ISME16 hairball that got me thinking. Conferences are a great place to catch up with old friends and make new ones - how does that play out on our Twitter mention graph? I decided to apply graph clustering analysis to start to try and get an angle on this with visualization.

```{r}
cw <- cluster_walktrap(mention_graph)
cw
```

I tried a few different clustering algorithms built into `igraph`, some were painfully slow, and some did a terrible job resolving modularity, but a weighted [random walk approach](http://arxiv.org/abs/physics/0512106) was super fast and turned out to do a great job with the default parameters. So our #ISME16 mention graph breaks out into 104 groups - what does that look like?

```{r}
V(mention_graph)$group <- membership(cw)
# Visualization is the same as before, except 
mention_graph_vn <- toVisNetworkData(mention_graph)
visNetwork(nodes = mention_graph_vn$nodes, edges = mention_graph_vn$edges,
           width = "100%", height = "800px") %>%
  visIgraphLayout(physics = T, smooth = T) %>%
  visEdges(arrows = "to") %>%
  visOptions(highlightNearest = TRUE)
```

Aside from being super pretty, I think this visualization is really neat and helps highlight the substructure in this highly interconnected graph. Many of those highly connected speakers form a big clique in the center of the graph and provide a lot of interconnectivity. But walk out from that and you can see some cool trends too. For example, labs and colleagues tend to form their own cliques (check out Kelly Wrighton's lab and other associated Ohio State people in particular). And there are people at the edges of those cliques that make the bridge to other groups and other cliques, really embracing the networking aspect of conferences (at least on Twitter!) Take a look and see where you landed. Does it feel about right? Does seeing the interactions this way change the way you think about using Twitter at conferences? Let me know!

### Wrapup

This is about as far as I got before I thought I should post this online somewhere. Thanks again for a great conference everyone and I hope to see you all in Leipzig for [#ISME17](https://twitter.com/hashtag/ISME17?src=hash)!